---
title: "Forecasting CMS Market Basket Using Discrete Time Series Methods (ARIMA)"
author: "Ayesha Parsotan"
date: '`r Sys.Date()`'
format:
  html:
    theme: litera
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
mainfont: "Times New Roman"
css: styles.css
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Presentation: [slides.html](slides.html){target="_blank"}

## Introduction

The Centers for Medicare and Medicaid Services (CMS) publishes market
basket indices to measure the input price inflation healthcare providers
can expect. These indices summarize expected changes in cost of labor,
equipment, supplies, etc. required to deliver care and they stand as the
standard for annual payment updates across major Medicare payment
systems. Since Market Basket updates directly influence reimbursement
levels, they play a central role in budgeting, actuarial modeling and
financial planning throughout the healthcare industry. Organizations
rely on these indices to anticipate future cost trends, evaluate
financial risk and align internal expectations with federal payment
policy @cms_hhpps_final_rule_2026.

Within this broader framework, the Home Health market basket is
particularly important. Under the Home Health Prospective Payment System
(HH PPS), the market basket determines the annual payment update for
home health agencies, affecting provider revenue, staffing capacity, and
operational strategy. Although the final Home Health market basket
update is published annually, CMS also releases quarterly projected
index levels, creating a long historical times series that is suitable
for statistical modeling.

Time‑series methods, particularly ARIMA models have been widely applied
in healthcare economics to forecast medical expenditures, inflation, and
service demand. Prior studies have demonstrated that ARIMA based
approaches perform well in modeling healthcare spending and medical
price behavior, including forecasting national health expenditures
@dritsakis2019time, providing broad overviews of health forecasting
methodologies @soyiri2013overview, and predicting medical service demand
using hybrid ARIMA‑based models @huang2020medical. ARIMA models have
also been used to forecast macroeconomic indicators in complex systems
@vafin2020forecasting, reinforcing their suitability for modeling
structured economic indices such as the CMS market baskets. Foundational
work on medical price indices further highlights the importance of
understanding the time‑series properties of healthcare cost measures
@berndt2000medical.

Forecasting the Home Health market basket can support financial
planning, inform strategic decision‑making, and help organizations
anticipate implications for cost‑of‑care management. This project
applies classical time‑series methods to model and forecast the CMS Home
Health Market Basket Index.

## Methods

A **Time Series** is a collection of observations $x_t$ made
sequentially in time $T_0$. A somewhat unique feature of time series and
their models is that we cannot assume that the observations arise
independently from a common population. Studying models that incorporate
dependence is the key concept in time series analysis @cryer_chan_2010.

In general, the goal is to emphasize plotting methods that are
appropriate and useful for finding patterns that will lead to suitable
models for our time series data. Finding an appropriate model allows us
to then forecast future values based on past-behavior.

There are three main steps in the process of finding an appropriate
model.

1.  Model Specification (Identification)
2.  Model Fitting
3.  Model Diagnostics

**Discrete-Time Series:** Observations are recorded at distinct, evenly
spaced time point. Examples include: annual temperature, quarterly
economic indicators.

**Continuous-Time Series:** Collection of observations that are made
continuously over some time interval. For example: network traffic
measured in real time.

The CMS Home Health Market Basket is observed quarterly, meaning the
data occurs at fixed, evenly spaced intervals. Since the time series
advances in discrete steps, all models evaluated in this study are
formulated within a discrete time series stochastic process framework.

### Stationarity and Time Series Foundations

To make statistical inferences about the structure of a stochastic
process on the basis of an observed record of that process, we must
usually make some simplifying assumptions about the structure. The most
important assumption is that of stationarity. The basic idea of
stationarity is that that probability laws that govern behavior of the
process do not change over time @cryer_chan_2010.

A process $\{Y_t\}$ is strictly stationary if the joint distribution of
$(Y_{t_1}, \ldots, Y_{t_n})$ is the same as the joint distribution of
$(Y_{t_1-h}, \ldots, Y_{t_n-h})$ for all choices of time points
$t_1,\ldots,t_n$ and all choices of time lag $h$.

$$
f(Y_{t_1}, \ldots, Y_{t_n}) = f(Y_{t_1-h}, \ldots, Y_{t_n-h})
$$

A process $\{Y_t\}$ is **weakly stationary** if it satisfies that
following conditions:

Mean is constant: $$ E(Y_t) = \mu $$

Variance is constant: $Var(Y_t) = \sigma^2$

Autocovariance depends only on lag: $\gamma_{t,t-k} = \gamma_k$

Autocorrelation: $\gamma_{t,t-k} = \rho_k$

Weak stationarity is sufficient for ARMA and ARIMA models because they
assume that the underlying statistical process is stable over time.
However, most data sets are **non-Stationary**: without constant mean
over time, and must be transformed.

$$
X_t = \mu_t + Y_t, \\
\ where \ E(Y_t) = 0
$$ $$
\mu_t \text{ nonconstant } \;\Longrightarrow\; X_t \text{ is non-stationary}
$$

After removing systematic components of a non-stationary series, the
remaining series should resemble **white noise**: a sequence of
independent, identically distributed random variables.

```{r}
wn <- rnorm(500)  
plot.ts(wn, main = "White Noise Series")



```

In practice, stationarity is assessed using visual diagnostics and
formal statistical tests before fitting time series models.

The **Augmented Dickey-Fuller (ADF)** test is used to assess
stationarity.

$$
H_0: \text{The series is NOT stationary} \quad\text{vs.}\quad H_1: \text{The series is stationary}
$$

If the *p*-value is greater than 0.05, we will fail to reject $H_0$ and
we can use trend or seasonal estimation/elimination methods, transform
the times series data set and fit the non-stationary models.

**Trend:** a trend is a long term increase or decrease in the level of a
time-series. It reflects persistent directional movement over time.
Trends violate stationarity because the mean of the series changes over
time. Differencing or de-trending is used to remove trend components.

A non-stationary trend implies: $E[x_t] \neq E[x_{t-k}]$

Trend differencing removes this component: $\nabla x_t = x_t - x_{t-1}$

**Periodicity** refers to any recurring pattern in the time series,
regardless whether the cycle length is fixed or known. It indicates that
the series contains repeating behavior, but the timing of the
repetitions may vary.

**Seasonality** is a specific, structured form of periodicity where the
cycle length is known, fixed and repeats at regular intervals.

$E[x_t] = E[x_{t-s}],\ \mathrm{Var}(x_t) = \mathrm{Var}(x_{t-s})$

All seasonal patterns are periodic, but not all periodic patterns are
seasonal. Seasonal structure can be removed or modeled using seasonal
differencing or SARIMA models, which incorporate seasonal autoregressive
and moving-average components.

**Heteroskedasticity** occurs when the variance of the series changes
over time. This violates the assumption of constant variance and can
distort model estimation.

```{r}
par(mfrow =c(2,2))
data("uspop")
plot(uspop, ylab = "", yaxt = "n", main = "Trend")

data("lynx")
plot(lynx, ylab = "", yaxt = "n", main = "Periodicity/Seasonality")

data("JohnsonJohnson")
plot(JohnsonJohnson, ylab="",  yaxt = "n", main = "Heteroskedasticity")

data("AirPassengers")
plot(AirPassengers, ylab="",  yaxt = "n", main = "Heteroskedasticity, Seasonality, Trend")
```

### Models for Stationary and Non-Stationary Time Series

Let $Y_t$ denote the observed times series, and $e_t$ represent an
unobserved white noise series. A **general linear process**, $Y_t$ is
one that can represented as a weighted linear combination of present and
past white noise terms as:

$$
Y_t = \mu + \sum_{j=0}^{\infty} \psi_j e_{t-j}
$$

where $\mu$ is a constant mean, $\{\psi_j\}$ is a sequence of
coefficients, and $\{e_t\}$ is a white-noise innovation process.

**Stationary Models**

A **Moving Average (MA)** process of order *(q)*, written MA(*q),*
models the current values of the series using the current error term and
a finite number of past error terms. The standard form is:

$$Y_t = \mu - \theta_1 e_{t-1} - \cdots - \theta_q e_{t-q} + e_t$$

An **Autoregressive (AR)** process of order *(p)*, written AR*(p)*,
expresses the current values of the series as a linear combination of
its past values.

$$Y_t = \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p} + e_t$$

An **Autoregressive Moving Average (ARMA)** model is appropriate for
stationary time series. It combines AR terms: dependence on past values
and MA terms: dependence on past shocks.

$$
Y_t
= \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p}
+ e_t
+ \theta_1 e_{t-1} + \cdots + \theta_q e_{t-q}.
$$

**Non-Stationary Models**

An **Autoregressive Integrated Moving Average (ARIMA)** model is an
extension of the ARMA model that incorporates non-seasonal differencing
or detrending to remove trend and achieve stationarity.

$$
\phi(B)(1-B)^d Y_t = \theta(B)e_t
$$

**Differencing** or **de-trending** transforms a non-stationary series
into a stationary one, allowing ARMA structure to be applied to the
differenced data. ARIMA models are appropriate when the series shows
trend but no seasonal pattern.

$$
\nabla^{d} X_t = (1 - B)^{d} X_t
$$

$$
\nabla X_t = X_t - X_{t-1}
$$

$$
\tilde{Y}_t = X_t - (\hat{\beta}_0 + \hat{\beta}_1 t)
$$

**Seasonal Autoregressive Integrated Moving Average** models extend the
ARIMA framework to account for seasonal dependence in a time series. A
SARIMA*(p,d,q)(P,D,Q)s* model includes both nonseasonal components
*(p,d,q)* and seasonal components *(P,D,Q)* with seasonal period *s*.

The model is defined by applying both nonseasonal differencing *(1−B)d*
and seasonal differencing *(1−Bs)D* to produce a stationary series.

A time series ${Yt}$ follows a SARIMA*(p,d,q)(P,D,Q)s* model if the
differenced series is a stationary ARMA *(p,q)* process with additional
seasonal AR and MA terms.

$$
\phi(B)\,\Phi(B^s)\,(1-B)^d (1-B^s)^D Y_t
=
\theta(B)\,\Theta(B^s)\, e_t
$$

### Model Specification (Identification)

In model specification, the classes of time series models are selected
that may be appropriate for a given observed series. A model chosen at
any time is tentative and subject to revision later on in the analysis.
In choosing a model, we should attempt to adhere to the **principle of
parsimony**; that is a model should require the smallest number of
parameters that will adequately represent the time series
@cryer_chan_2010.

The **Autocorrelation Function (ACF)** measures the correlation between
$x_t$ and its lagged values $x_t-{_k}$ :

$$
\rho_k = \mathrm{Corr}(x_t, x_{t-k})
$$

A sharp cutoff after lag q suggests an MA *(q)* component. A gradual,
exponential decay suggests an AR or ARMA structure. Seasonal spikes at
multiples of the seasonal period *s* indicate seasonal MA terms.

The **Partial Autocorrelation Function (PACF)** measures the correlation
between $x_t$ and $x_t-{_k}$ after removing the effects of intermediate
lags:

$$\phi_{kk} = \mathrm{Corr}(x_t, x_{t-k} \mid x_{t-1}, \ldots, x_{t-k+1})$$

A sharp cutoff after lag *p* suggests an AR*(p)* component. A gradual
decay suggests an MA or ARMA structure. Seasonal PACF spikes at lag s
indicate seasonal AR terms. A PACF is especially useful for identifying
autoregressive behavior.

The **Extended Autocorrelation Function (EACF)** provides a systematic
way to identify ARMA *(p,q)* orders when ACF and PACF patterns are
ambiguous. The EACF constructs a grid where Rows = AR orders *p*,
Columns = MA orders *q*. Each cell is marked as: "o" (model is
plausible), "x" (model is not plausible). The grid is scanned for the
first clean triangle of "o", and the top left corner of the triangle
gives ARMA*(p,q)* model. The EACF is especially valuable when ACF and
PACF do not show clear cutoffs and the series has mixed AR and MA
behavior.

```{r}
ma1.1.s=arima.sim(list(order=c(0,0,1),ma=-0.9),100)
ma2.1.s=arima.sim(list(order=c(0,0,2),ma=c(-1,0.6)),100)
ar1.1.s=arima.sim(list(order=c(1,0,0),ar=0.9),100)
ar2.1.s=arima.sim(list(order=c(2,0,0),ar=c(1.5,-0.75)),100)
arma1.1.s=arima.sim(list(order=c(1,0,1),ar=0.6,ma=-0.3),100)

library(TSA)
par(mfrow = c(2, 3), oma = c(0, 0, 3, 0), mar = c(4, 4, 4, 1))
acf(ma1.1.s, main = "MA(1)")
acf(ma2.1.s, main = "MA(2)")
acf(ar1.1.s, main = "AR(1)")
pacf(ar1.1.s, main = "PACF AR(1)")
acf(ar2.1.s, main = "AR(2)")
pacf(ar2.1.s, main = "PACF AR(2)")


eacf(arma1.1.s)
```

After candidate models are manually identified, they are compared using
the **Corrected Akaike Information Criterion (AICc):**

$$
AIC_c = AIC + \frac{2k(k+1)}{n - k - 1}
$$

where:

$$
AIC = -2\log(L) + 2k
$$

$L$ is the maximized likelihood,\
$k$ is the number of estimated parameters, and\
$n$ is the sample size.

AICc adjusts for small sample sized and prevents overfitting. A lower
AICc equates to a better models, differences greater than 2 indicate
meaningful improvement and the most adequate model is selected.

In addition to manual identification using ACF, PACF, EACF, this study
uses **auto.arima()** function from the `forecast` package to support
model selection. The auto.arima() function automatically determines,
*p,d,q* and seasonal orders. It searches across candidate ARIMA/SARIMA
models and selects the models with the lowest AICc. This function
provides an objective benchmark, confirms or refines manually identified
models and ensures the final model is optimal under AICc.

### Model Fitting (Parameter Estimation)

Once the model orders *(p,d,q)* have been identified, the next step is
to estimate the parameters of the autoregressive (AR) , moving average
(MA), and combined ARMA models. Parameter estimation determines the
values of the AR co-effecient $\phi_i$ , the MA co-efficient $\theta_i$
and the innovation variance $\sigma^2$ . Different estimation methods
apply depending on the model class.

1.  *Yule-Walker Estimation \| (AR)* : relate the AR coefficients to the
    autocovariance function.

    $$
    \gamma_k = \phi_1 \gamma_{k-1} + \cdots + \phi_p \gamma_{k-p}, 
    \qquad k = 1,\ldots,p.
    $$

2.  *Burg Method \| (AR)* : estimates AR parameters by minimizing
    forward and backward prediction errors while enforcing stability.

3.  *Conditional Sum of Squares (CSS*) \| *(AR, MA, ARMA)* : estimates
    AR and MA parameters by minimizing the sum of squared residuals. It
    is useful for initial parameters but less efficient than MLE

    $$
        \sum_{t = p+1}^{T} \varepsilon_t^2
    $$

4.  *Maximum Likelihood Estimation (MLE) \| (AR)* : maximizes the
    Gaussian likelihood, yielding efficient and asymptotically normal
    estimates. It is the preferred method for MA models because it
    accounts for the dependence on unobserved errors. This is required
    for the AICc model comparison and the model must pass the normality
    assumption for this estimation.

    $$
    L(\theta \mid y_1,\ldots,y_T),
    $$

5.  *Innovations Algorithm \| (MA)* : recursively computes optimal
    linear predictors and innovation variances, allowing estimation of
    MA parameters. It is efficient for pure MA models and handles
    missing data well.

6.  *Hannan-Rissanen Method \| (MA, ARMA)* : provides initial estimates
    for MA parameters, fits a higher order AR model to obtain
    pre-liminary residuals. This method regresses $Y_t$ on lagged values
    and estimated residuals and uses these as starting values for MLE.
    This method is widely used for ARMA models.

    |  |  |  |
    |------------------------|------------------------|------------------------|
    | **Model Type** | **Estimation Methods** | **Notes** |
    | AR *(p)* | Yule-Walker, Burg, CSS, MLE | Yule-Walker/Burg for initial estimates; MLE for refinement |
    | MA *(q)* | Innovations, Hannan-Rissanen, CSS, MLE | MA requires numerical methods; Hannan-Rissanen gives good starting values |
    | ARMA *(p,q)* | Hannan-Rissanen, CSS, MLE | Hannan-Risannen and MLE are standard |
    | ARIMA *(p,d,q)* | CSS, MLE |  |
    | SARIMA *(p,d,q)(P,D,Q)s* | CSS, MLE |  |

    : *Summary of Parameter Estimation Methods*

### Model Diagnostics

Model diagnostics focus on assessing how well a model fits the observed
data and if the fit is inadequate, identifying suitable modifications.
The main objective is to evaluate the adequacy of the selected model.
After fitting the time series model, diagnostic checks on the residuals
should be performed to evaluate the model's fit. Residuals should behave
like white noise and should exhibit no autocorrelation, constant
variance, zero mean, approximate normality and no remaining seasonal
structure.

**Statistical Inference for Model Coefficients**

Each estimated parameter is evaluated for statistical significance. For
a generic coefficient $\beta$, the hypothesis test is:

$$
H_0:\ \beta = 0 \quad\text{(coefficient is not significant)}
$$

$$
H_1:\ \beta \neq 0 \quad\text{(coefficient is significant)}
$$

the test statistic is:

$$
z = \frac{\hat{\beta}}{SE(\hat{\beta})}
$$

and the corresponding p-value determines whether the coefficient is
significant. If the *p*-value is less than 0.05, we reject $H_0$ and the
coefficient is significant. Significant testing helps to identify
parameter redundancy and prevents overfitting by ensuring that only
meaningful terms are retained.

**Standardized Residuals:**

$$
e_t^{*} = \frac{\hat{e}_t}{\hat{\sigma}}
$$

residuals are examined to assess whether the model errors exhibit: zero
mean, constant variance, no visible patterns, no remaining seasonality.
A well fitted model produces residuals that fluctuate randomly around
zero with no systematic structure.

**Normality Assessment:**

Approximate normality improves the reliability of forecast intervals.
Normality was evaluated using Normal Q-Q plot, Shapiro-Wilk test and a
histogram of standardized residuals. Extreme skewness or heavy tails may
indicate the need for further transformation or alternative modeling.

**Residual ACF and PACF:**

The residual ACF was examined to detect remaining serial dependence.
Significant spikes would indicate that the model failed to capture all
AR or MA structure. For an adequate model, all residual autocorrelations
lies within the 95% CI bounds, no seasonal spikes occur at lag and there
is no slow decay or cyclical pattern.

The residual PACF of the residuals are also inspected. Significant
spikes in the PACF suggest remaining autoregressive behavior. A well
specified model should show no significant partial autocorrelations and
no seasonal partial autocorrelations.

Together, the residual ACF and PACF confirm whether the model has
removed all linear dependence.

**Ljung-Box Test:**

The Ljung-Box Test formally evaluates whether the residuals are
uncorrelated:

$$
\begin{aligned}
H_0:\ &\text{Residuals are white noise} \\
H_1:\ &\text{Residuals are not white noise}
\end{aligned}
$$

A large *p*-value greater than 0.05 indicates failure to reject $H_0$
supporting the adequacy of the model.

### Forecasting

Forecasting in AR, MA, ARMA, ARIMA, SARIMA models is based on projecting
the conditional expectation of future values given all information
available up to time $T$:

This framework applies uniformly across all stationary and
non-stationary models discussed. The goal of forecasting is to estimate
future values of the series $Y_{T+h}$ for $h \ge 1$

while quantifying the uncertainty associated with those estimates.

The terms predicting and forecasting are often used interchangeable,
however they have distinct meanings in time-series analysis. Prediction
refers to estimating an unknown value within the observed sample and
does not require temporal ordering. Forecasting refers specifically to
estimating values beyond the end of the observed sample, using temporal
dependence structure of the series.

**Forecast Accuracy Metrics: RMSE, MAPE, and MSE**

To evaluate the quality of the forecasts, several standard accuracy
metrics are computed from the forecast errors $e_t(\ell)$:

**Mean Squared Error (MSE)**

MSE measures the average squared deviation between the forecast and the
true value:

$$
\text{MSE} = \frac{1}{n} \sum_{t=1}^{n} e_t(\ell)^2
$$

It penalizes larger errors more heavily and is widely used in model
comparison.

**Root Mean Squared Error (RMSE)**

RMSE is the square root of MSE:

$$
\text{RMSE} = \sqrt{\text{MSE}}
$$

RMSE is expressed in the same units as the original series, making it
easier to interpret. Lower RMSE values indicate tighter, more accurate
forecasts.

**Mean Absolute Percentage Error (MAPE)**

MAPE expresses forecast accuracy as a percentage:

$$
\text{MAPE} = \frac{100}{n} \sum_{t=1}^{n} \left| \frac{e_t(\ell)}{Y_{t+\ell}} \right|
$$

MAPE is scale‑free and intuitive, though it can be unstable when actual
values are near zero. Together, MSE, RMSE, and MAPE provide a
comprehensive assessment of forecast performance.

| Metric | Description |
|----------------------------|--------------------------------------------|
| MSE | Mean Squared Error: average of squared forecast errors. |
| RMSE | Root Mean Squared Error: square root of MSE; same units as the data. |
| MAPE | Mean Absolute Percentage Error: average absolute percent error. |

**Optional Training–Testing Approach**

Although the primary analysis in this study relies on full sample
estimation of the ARIMA model, an optional evaluation strategy involves
partitioning the data into a training set and a test set. In this
approach, the model is estimated using the training portion of the
series, and its out of sample performance is assessed by forecasting the
remaining observations. This method provides an additional way to
evaluate forecast accuracy, but it is not required for the ARIMA
estimation itself. When used, the same accuracy metrics MSE, RMSE, and
MAPE are computed on the test set to quantify out of sample performance.

## Analysis and Results

### Data Preparation and Exploration

```{r, warning=FALSE, echo=T, message=FALSE}
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(readr)
library(dplyr)
library(zoo)
library(tseries)
library(TSA)
library(forecast)
library(broom)
library(lmtest)
library(itsmr)
```

| Category | Package | Purpose |
|------------------|------------------|-----------------------------------|
| *Data Import & Wrangling* | `tidyverse` | Unified ecosystem for data manipulation, cleaning, transformation, and piping. |
|  | `dplyr` | Grammar of data manipulation: filter, mutate, summarize, join, group operations. |
|  | `readr` | Fast import of rectangular data (CSV, TSV, delimited files). |
|  | `zoo` | Infrastructure for irregular time series; rolling operations and index handling. |
|  | `dslabs` | Provides datasets and helper functions for data science workflows. |
|  | `broom` | Converts model objects into tidy data frames for analysis and reporting. |
| *Visualization & Reporting* | `ggthemes` | Additional ggplot2 themes for clean academic and professional styling. |
|  | `ggrepel` | Prevents overlapping text labels in ggplot2 visualizations. |
|  | `knitr` | Integrates R code and output into Quarto/R Markdown documents. |
| *Time-Series Modeling* | `tseries` | Classical time-series tools: ADF test, KPSS, GARCH, ARIMA utilities. |
|  | `TSA` | Shumway & Stoffer time-series functions: periodograms, filtering, ARMA tools. |
|  | `forecast` | Forecasting framework: ARIMA, ETS, STL, accuracy metrics, forecast plots. |
| *Statistical Testing & Diagnostics* | `lmtest` | Diagnostic tests for regression and time-series models (Ljung–Box). |
|  | `itsmr` | Time-series diagnostics: turning points, rank tests, residual randomness tests. |

The raw CMS Market Basket Index dataset was imported and screened to
remove incomplete observations by excluding rows with missing or blank
values in the HH Index Levels field. The index variable was then changed
to numeric to ensure proper handling in subsequent time‑series
operations. Column names were standardized by renaming the market basket
date field to Date and the index measure to Index for clarity and
consistency. The dataset was then sorted chronologically to preserve the
temporal ordering required for time‑series modeling. Finally, the
cleaned and ordered Index series was transformed into a quarterly
time‑series object (ts) with a start date of 2010 Q1, producing a
structured dataset suitable for ARIMA‑based analysis.

```{r}
# Import Data Set
Data_Import <- read.csv("C:/Users/ayesh/OneDrive/Desktop/MS Data Science/IDC6940 Projects in Data Science/IDC6940_ARIMA/2025Q3 CMS HH MBI.csv")


# Remove N/A and Blank Values
Data <- Data_Import %>%
  filter(
    !is.na(`HH.Index.Levels`),          
    `HH.Index.Levels` != ""           
  )

# Convert HH Index Levels to Numeric
Data <- Data %>%
  mutate(`HH.Index.Levels` = as.numeric(`HH.Index.Levels`))

# Rename Headers
Data <- Data %>%
  rename(
    Date = `Market.Basket`,
    Index = `HH.Index.Levels`
  )

# Convert Date to Quarterly Format
Data <- Data %>%
  mutate(`Date` = as.yearqtr(Date, format = "%Y Q%q"))

# Sort Chronologically 
Data <- Data %>%
  arrange(`Date`)

# Convert to Time Series
TS_Data <- ts(Data$Index, frequency = 1)

```

### Preliminary Analysis

```{r}
library(dplyr)

Table1 <- Data %>%
  summarise(
    N = n(),
    Mean_Index = round(mean(Index, na.rm = TRUE),3),
    SD_Index = round(sd(Index, na.rm = TRUE),3),
    Min_Index = min(Index, na.rm = TRUE),
    Max_Index = max(Index, na.rm = TRUE),
    Start_Date = first(Date),
    End_Date = last(Date)
  )

knitr::kable(Table1, caption = "Table 1. Descriptive Statistics for CMS Home Health Market Basket Index (2010–2025)")
```

Table 1 provides a concise overview of the CMS Home Health Market Basket
Index from 2010 through 2025. Across the 62 quarterly observations, the
index exhibits a mean value of approximately 0.936 with a standard
deviation of 0.117, indicating modest variability over the study period.
The minimum observed value (0.789) and maximum (1.196) suggest a gradual
upward movement consistent with long‑term cost inflation in the home
health sector. The dataset spans from 2010 Q1 to 2025 Q2, ensuring
adequate temporal coverage for trend identification and subsequent
time‑series modeling. These summary statistics confirm that the series
is suitable for ARIMA‑based forecasting.

```{r, warning=FALSE, echo=TRUE, fig.width=12, fig.height=4}
data(TS_Data)
head(TS_Data)
summary (TS_Data)

par(mfrow = c(1, 2))
plot(TS_Data,
     main = "CMS Home Health Market Basket Index 2010–2025",
     xlab = "Date",
     ylab = "Index")

qqnorm(TS_Data)
qqline(TS_Data, col= "red")
```

The time series plot of the data indicates a clear upward trend with no
apparent seasonal component or change in variance over time. The Q-Q
plot shows that the CMS Home Health Market Basket data deviates from
normality. Moreover, the Shapiro-Wilk test is performed to test the
following hypothesis for normality.

$$
H_0: \text{The CMS HH Market Basket Index follows a Normal Distribution} \\ \quad\text{vs.} \\ \quad H_1: \text{The CMS HH Market Basket Index does NOT follow a Normal Distribution}
$$

```{r}
tidy(shapiro.test(TS_Data))
```

*p*-value \< 0.05, thus we can reject the null hypothesis. We have
sufficient evidence to conclude that the CMS Home Health Market Basket
Data does not follow a normal distribution.

### Stationarity Test and Model Specification

ADF test is performed to test stationarity:

$$
H_0: \text{The CMS HH Market Basket Index is NOT stationary} \\ \quad\text{vs.} \\ \quad H_1: \text{The CMS HH Market Basket Index is stationary}
$$

```{r}
adf.test(TS_Data)
plot(diff(diff(TS_Data)))
```

Since the *p*-value \> 0.05, we fail to reject the null hypothesis. This
indicates that the CMS HH Market Basket Index is non-stationary in its
current form.

```{r}
par(mfrow = c(1,2))

acf(as.numeric(TS_Data))
pacf(as.numeric(TS_Data))
eacf(as.numeric(TS_Data))

tidy(auto.arima(TS_Data))
```

The ARIMA(1,2,0) model was selected as the most appropriate
specification based on the combined evidence from the ACF, PACF, and
EACF diagnostics. The raw series exhibited a clear upward trend, and the
slow decay in the ACF indicated non‑stationarity, supporting the need
for differencing. The PACF showed a dominant spike at lag 1 with
subsequent lags falling within the confidence bounds, consistent with an
AR(1) structure.The auto.arima() function indicated a differencing of 2
and estimated co-efficent of -0.5844 using MLE. Together, these
diagnostics pointed toward a parsimonious ARIMA(1,2,0) model that
adequately captures the underlying dependence structure of the series.

**Candidate Model**: ARIMA (1,2,0) - First order autoregressive process
with an integration of second order differencing. $$
(1 + 0.5844B)(1 - B)^2 Y_t = e_t
$$

### Modeling and Results

```{r}
fit_ARIMA120 <- arima(
  TS_Data,
  order = c(1, 2, 0)
)

coeftest(fit_ARIMA120)

fit_ARIMA120_MLE <- arima(TS_Data, order = c(1, 2, 0), method = "ML")


fit_ARIMA120_CSS <- arima(TS_Data, order = c(1, 2, 0), method = "CSS")

model.comparison.arima <- matrix(NA, ncol = 4, nrow = 1)
colnames(model.comparison.arima) <- c("MLE_Est", "MLE_SE", "CSS_Est", "CSS_SE")
rownames(model.comparison.arima) <- c("phi")

model.comparison.arima[1, ] <- c(
  fit_ARIMA120_MLE$coef[1],
  sqrt(diag(fit_ARIMA120_MLE$var.coef))[1],
  fit_ARIMA120_CSS$coef[1],
  sqrt(diag(fit_ARIMA120_CSS$var.coef))[1]
)

knitr::kable(
  model.comparison.arima,
  digits = 4,
  caption = "Comparison of AR(1) Estimates and Standard Errors for MLE and CSS ARIMA(1,2,0) Models"
)


```

The ARIMA(1,2,0) model parameters were estimated using both Maximum
Likelihood Estimation (MLE) and Conditional Sum of Squares (CSS) to
assess the stability of the autoregressive parameter. Both estimation
methods produced nearly identical AR(1) coefficients (–0.5844 for ML and
–0.5932 for CSS), with MLE standard error being smaller. The
auto.arima() function returned the ML estimate and The z‑test for the
AR(1) parameter shows that the coefficient is highly statistically
significant. This consistency across estimation methods strengthens
confidence in the ARIMA(1,2,0) specification as an appropriate model for
the data.

```{r}
res <- residuals(fit_ARIMA120)
test(res)
shapiro.test(res)
```

$$
H_0: \text{The Residuals are normal} \\ \quad\text{vs.} \\ \quad H_1: \text{The residuals are NOT normal}
$$

All residual diagnostics support the adequacy of the ARIMA(1,2,0) model.
The Ljung–Box test indicate no remaining autocorrelation. Nonparametric
tests (turning points, difference signs, and rank tests) confirm that
the residuals behave as iid noise.

The Shapiro–Wilk test shows no evidence against normality.The *p*-value
\> 0.5, so we fail to reject the null hypothesis, normality is not a
concern for the candidate model.

Together, these results demonstrate that the ARIMA(1,2,0) model provides
an appropriate fit for the CMS Home Health Market Basket Index.

```{r}
predict_ARIMA120=predict(fit_ARIMA120,n.ahead=8)
predict_ARIMA120$pred
predict_ARIMA120$se

upper_95 <- predict_ARIMA120$pred + 1.96 * predict_ARIMA120$se
lower_95 <- predict_ARIMA120$pred - 1.96 * predict_ARIMA120$se

plot(TS_Data, xlim = c(1, length(TS_Data) + 8),
     ylim = range(c(TS_Data, upper_95, lower_95)),
     main = "ARIMA(1,2,0) Forecast",
     ylab = "Index", xlab = "Time")

lines((length(TS_Data)+1):(length(TS_Data)+8), predict_ARIMA120$pred, col = "blue", type="b", pch=1)

lines((length(TS_Data)+1):(length(TS_Data)+8), upper_95, col = "red", lty = 2)
lines((length(TS_Data)+1):(length(TS_Data)+8), lower_95, col = "red", lty = 2)
```

```{r}
library(forecast)
library(knitr)

fitted_vals <- fitted(fit_ARIMA120)

accuracy_results <- data.frame(
  MAE  = mean(abs(residuals(fit_ARIMA120))),
  RMSE = sqrt(mean(residuals(fit_ARIMA120)^2)),
  MAPE = mean(abs(residuals(fit_ARIMA120) / TS_Data)) * 100,
  MSE  = mean(residuals(fit_ARIMA120)^2)
)


kable(
  accuracy_results,
  digits = 4,
  caption = "In-Sample Forecast Accuracy Metrics for ARIMA(1,2,0) Model"
)

```

The ARIMA(1,2,0) model demonstrated excellent in-sample forecasting
performance across all evaluation metrics. The mean absolute error
(MAE=0.002) and the root mean squared error (RMSE = 0.0024), indicate
that prediction errors are minimal and consistent. Additonally, the mean
absolute percentage error (MAPE=0.2078%) suggesting that the forecast
deviates from actual values by less than 1% on average, reflecting a
very high level of accuracy. Overall, these results indicate that the
model captures the underlying patterns in the data effectively.

### Conclusion

This study applied a classical time‑series modeling framework to
evaluate and forecast the CMS Home Health Market Basket Index, a key
economic indicator used to adjust Medicare reimbursement rates. After
exploring the structure of the series, the ACF, PACF, and EACF
diagnostics consistently indicated the presence of a strong
deterministic trend and supported the need for second‑order
differencing. Model identification procedures converged on an
ARIMA(1,2,0) specification, which was subsequently estimated using
maximum likelihood. The autoregressive parameter was statistically
significant and stable across estimation methods. Comprehensive residual
diagnostics including tests for autocorrelation, normality, and
randomness confirmed that the residuals behaved as white noise,
indicating that the model adequately captured the underlying dynamics of
the series. Forecast accuracy metrics (MAE, RMSE, MAPE, MSE) further
supported the model’s suitability by showing that the in sample errors
were small and consistent with a well‑fitting ARIMA process. Taken
together, these results validate the ARIMA(1,2,0) model as an
appropriate and reliable tool for short‑term forecasting of the Home
Health Market Basket Index, providing a sound basis for anticipating
future cost‑inflation adjustments within the home health sector.

## References
