---
title: "Forecasting CMS Market Basket Using Discrete Time Series Methods (ARIMA)"
author: "Ayesha Parsotan"
date: '`r Sys.Date()`'
format:
  html:
    theme: litera
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
mainfont: "Arial"
css: styles.css
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Presentation: [slides.html](slides.html){target="_blank"}

## Introduction

The Centers for Medicare and Medicaid Services (CMS) publishes market
basket indices to measure the input price inflation healthcare providers
can expect. These indices summarize expected changes in cost of labor,
equipment, supplies, etc. required to deliver care and serve as the
standard for annual payment updates across major Medicare payment
systems. Since Market Basket updates directly influence reimbursement
levels, they play a central role in budgeting, actuarial modeling and
financial planning throughout the healthcare industry. Organizations
rely on these indices to anticipate future cost trends, evaluate
financial risk and align internal expectations with federal payment
policy.

Within this broader framework, the Home Health market basket is
particularly important. Under the Home Health Prospective Payment System
(HH PPS), the market basket determines the annual payment update for
home health agencies, affecting provider revenue, staffing capacity, and
operational strategy.

The final Home Health market basket is published annually, but projected
estimates are published quarterly and forms a long historical time
series, making it suitable for statistical modeling. Time series
methods, including ARIMA models have been widely applied in healthcare
expenditure forecasting and have demonstrated strong performance in
capturing inflationary and utilization trends (reference). Forecasting
the Home Health market basket can support financial planning, inform
strategic decision making and help organizations anticipate implications
for cost of care management.

This project applies time-series methods to model and forecast the CMS
market basket index which is a variable in computing market basket.

## Methods

A **Time Series** is a collection of observations $x_t$ made
sequentially in time $T_0$, and can be classified into two types:

**Discrete-Time Series:** Observations are recorded at distinct, evenly
spaced time point. Examples include: annual temperature, quarterly
economic indicators. $T_0 = {1,2,3,...T}$

**Continuous-Time Series:** Collection of observations that are made
continuously over some time interval. For example: network traffic
measured in real time. Continuous time-series models require
differential equations and are not used in ARIMA modeling.

The CMS Home Health Market Basket is observed quarterly, meaning the
data occurs at fixed, evenly spaced intervals. Since the time series
advances in discrete steps, all models evaluated in this study are
formulated within a discrete time series stochastic process framework.

The objective of Times Series modeling is to not only provide a model of
the data but to forecast future values based on past-behavior. The first
goal in time series modeling is to reduce the analysis needed to a
simpler case: Eliminate Trend, Periodicity or Seasonality and
Heteroskedasticity. Thereafter, we can model the remainder or residuals
as dependent but Identically distributed.

### Stationarity and Time Series Foundations

A process $\{Y_t\}$ is strictly stationary if the joint distribution of
$(Y_{t_1}, \ldots, Y_{t_n})$ is the same as the joint distribution of
$(Y_{t_1-h}, \ldots, Y_{t_n-h})$ for all choices of time points
$t_1,\ldots,t_n$ and all choices of time lag $h$.

$$
f(Y_{t_1}, \ldots, Y_{t_n}) = f(Y_{t_1-h}, \ldots, Y_{t_n-h})
$$

A process $\{Y_t\}$ is **weakly stationary** if:

Mean is constant: $E(Y_t) = \mu$

Variance is constant: $Var(Y_t) = \sigma^2$

Autocovariance depends only on lag: $\gamma_{t,t-k} = \gamma_k$

Correlation:\_\_\_\_\_

Weak Stationarity is sufficient for ARMA models assume that the
underlying statistical process is stable over time. Non-Stationary
series must be transformed, typically through the process of
differencing or seasonal differencing before ARIMA/SARIMA modeling.

In practice, we always tests stationarity of time series data.

The **Augmented Dickey-Fuller (ADF)** test is used to assess
stationarity.

$$
H_0: \text{The series is NOT stationary} \quad\text{vs.}\quad H_1: \text{The series is stationary}
$$

If the p-value is above 0.05, we will fail to reject $H_0$ and we can
use trend or seasonal estimation/elimination methods, transform the
times series data set and Fit the non-stationary models.

**Trend:** a trend is a long term increase or decrease in the level of a
time-series. It reflects persistent directional movement over time.
Trends violate stationarity because the mean of the series changes over
time. Differencing or de-trending is used to remove trend components.

A non-stationary trend implies: $E[x_t] \neq E[x_{t-k}]$

Trend differencing removes this component: $\nabla x_t = x_t - x_{t-1}$

```{r}
data("uspop")
plot(uspop)
```

**Periodicity** refers to any recurring pattern in the time series,
regardless whether the cycle length is fixed or known. It indicates that
the series contains repeating behavior, but the timing of the
repetitions may vary.

**Seasonality** is a specific, structured form of periodicity where the
cycle length is known, fixed and repeats at regular intervals.

$E[x_t] = E[x_{t-s}],\ \mathrm{Var}(x_t) = \mathrm{Var}(x_{t-s})$

All seasonal patterns are periodic, but not all periodic patterns are
seasonal. Seasonal differencing or SARIMA models are used to remove or
model seasonal structure.

```{r}
data("lynx")
plot(lynx)
```

**Heteroskedasticity** occurs when the variance of the series changes
over time. This violates the assumption of constant variance and can
distort model estimation. Specialized models, for example: GARCH are
used to eliminated variance.

```{r}
data("JohnsonJohnson")
plot(JohnsonJohnson)

data("AirPassengers")
plot(AirPassengers)
```

include snippet about IID noise and White NOISE

### AR (p), MA (q), ARMA (p,q), ARIMA (p,d,q) SARIMA

**Stationary Models**

A **Moving Average (MA)** process of order (q), written MA(q), models
the current values of the series using the current error term and a
finite number of past error terms. The standard form is:

$Y_t = \mu - \theta_1 e_{t-1} - \theta_2 e_{t-2} - \cdots - \theta_q e_{t-q} + e_t$

An **Autoregressive (AR)** process of order (p), written AR(p),
expresses the current values of the series as a linear combination of
its past values.

$y_t = \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p} + e_t$

An **Autoregressive Moving Average (ARMA)** model is appropriate for
stationary time series. It combines AR terms: dependence on past values
and MA terms: dependence on past shocks.

$$
y_t
= \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p}
+ \varepsilon_t
+ \theta_1 \varepsilon_{t-1} + \cdots + \theta_q \varepsilon_{t-q}.
$$

**Non-Stationary Models**

An **Autoregressive Integrated Moving Average (ARIMA)** model is an
extension of the ARMA model that incorporatez non-seasonal differencing
to remove trend and achieve stationarity.

$$
\Phi(B)\,(1 - B)^{d}\, y_t \;=\; \Theta(B)\,\varepsilon_t
$$

where:

p: number of AR terms

d: number of differences

q: number of MA terms

B: backshift operator.

Differencing transforms a non-stationary series into a stationary one,
allowing ARMA structure to be applied to the differenced data. ARIMA
models are appropriate when the series shows trend but no seasonal
pattern.

A **Seasonal Autoregressive Integrated Moving Average (SARIMA)** is an
extension of ARIMA that incorporates seasonal components.

$$
\Phi(B^{s})\,\phi(B)\,(1 - B)^{d}\,(1 - B^{s})^{D}\, y_t
\;=\;
\Theta(B^{s})\,\theta(B)\,\varepsilon_t .
$$

$$
\begin{aligned}
\phi(B) &= \text{nonseasonal AR polynomial}, \\
\theta(B) &= \text{nonseasonal MA polynomial}, \\
\Phi(B^{s}) &= \text{seasonal AR polynomial}, \\
\Theta(B^{s}) &= \text{seasonal MA polynomial}, \\
(1 - B)^{d} &= \text{nonseasonal differencing operator}, \\
(1 - B^{s})^{D} &= \text{seasonal differencing operator}
\end{aligned}
$$

It is written as:

SARIMA (p,d,q)(P,D,Q)s

where:

(p,d,q): non-seasonal ARIMA orders

(P,D,Q) seasonal ARIMA orders

s: seasonal period

### Model Specification

There are three main steps in the process of finding the appropriate
model for a time-series: model identification, model fitting and model
diagnostics.

The **Autocorrelation Function (ACF)** measures the correlation between
$x_t$ and its lagged values $x_t-{_k}$ :

$\rho_k = \mathrm{Corr}(x_t, x_{t-k})$

A sharp cutoff after lag q suggests an MA (q) component. A gradual,
exponential decay suggests an AR or ARMA structure. Seasonal spikes at
multiples of the seasonal period s indicate seasonal MA terms.

The **Partial Autocorrelation Function (PACF)** measures the correlation
between $x_t$ and $x_t-{_k}$ after removing the effects of intermediate
lags:

$\phi_{kk} = \mathrm{Corr}(x_t, x_{t-k} \mid x_{t-1}, \ldots, x_{t-k+1})$

A sharp cutoff after lag p suggests an AR(p) component. A gradual decay
suggests an MA or ARMA structure. Seasonal PACF spikes at lag s indicate
seasonal AR terms. A PACF is especially useful for identifying
autoregressive behavior.

The **Extended Autocorrelation Function (EACF)** provides a systematic
way to identify ARMA (p,q) orders when ACF and PACF patterns are
ambiguous. The EACF constructs a grid where Rows = AR orders p, Columns
= MA orders q. Each cell is marked as: "o" (model is plausible), "x"
(model is not plausible). The grid is scanned for the first clean
triangle of "o", and the top left corner of the triangle gives ARMA(p,q)
model. The EACF is especially valuable when ACF and PACF do not show
clear cutoffs and the series has mixed AR and MA behavior.

```{r}
#Simulations
ma1.1.s=arima.sim(list(order=c(0,0,1),ma=-0.9),100)
ma2.1.s=arima.sim(list(order=c(0,0,2),ma=c(-1,0.6)),100)
ar1.1.s=arima.sim(list(order=c(1,0,0),ar=0.9),100)
ar2.1.s=arima.sim(list(order=c(2,0,0),ar=c(1.5,-0.75)),100)
arma1.1.s=arima.sim(list(order=c(1,0,1),ar=0.6,ma=-0.3),100)

library(TSA)
par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))
acf(ma1.1.s, main = "MA(1)")
acf(ma2.1.s, main = "MA(2)")
acf(ar1.1.s, main = "AR(1)")
pacf(ar1.1.s, main = "PACF AR(1)")
acf(ar2.1.s, main = "AR(2)")
pacf(ar2.1.s, main = "PACF AR(2)")


eacf(arma1.1.s)
```

After candidate models are manually identified, they are compared using
the **Corrected Akaike Information Criterion (AICc):**

$$
AIC_c = AIC + \frac{2k(k+1)}{n - k - 1}
$$

where:

$$
AIC = -2\log(L) + 2k
$$

$L$ is the maximized likelihood,\
$k$ is the number of estimated parameters, and\
$n$ is the sample size.

AICc adjusts for small sample sized and prevents overfitting. A lower
AICc equates to a better models, differences greater than 2 indicate
meaningful improvement and the most adequate model is selected.

In addition to manual identification using ACF, PACF, EACF, this study
uses **auto.arima()** function from the *forecast* package to support
model selection. The auto.arima() function automatically determines,
p,d,q and seasonal orders. it uses unit root tests such as ADF and KPSS
to choose differencing. It searches across candidate ARIMA/SARIMA models
and selects the models with the lowest AICc. This function provides an
objective benchmark, confirms or refines manually identified models and
ensures the final model is optimal under AICc.

### Parameter Estimation

Once the model orders (p,d,q) have been identified, the next step is to
estimate the parameters of the autoregressive (AR) , moving average
(MA), and combined ARMA models. Parameter estimation determines the
values of the AR co-effecient $\phi_i$ , the MA co-efficient $\theta_i$
and the innovation variance $\sigma^2$ . Different estimation methods
apply depending on the model class.

1.  *Yule-Walker Estimation \| (AR)* : relate the AR coefficients to the
    autocovariance function.

    $$
    \gamma_k = \phi_1 \gamma_{k-1} + \cdots + \phi_p \gamma_{k-p}, 
    \qquad k = 1,\ldots,p.
    $$

2.  *Burg Method \| (AR)* : estimates AR parameters by minimizing
    forward and backward prediction errors while enforcing stability.

3.  *Conditional Sum of Squares (CSS*) \| *(AR, MA, ARMA)* : estimates
    AR and MA parameters by minimizing the sum of squared residuals. It
    is useful for initial parameters but less efficient than MLE

    $$
        \sum_{t = p+1}^{T} \varepsilon_t^2
    $$

4.  *Maximum Likelihood Estimation (MLE) \| (AR)* : maximizes the
    Gaussian likelihood, yielding efficient and asymptotically normal
    estimates. It is the preferred method for MA models because it
    accounts for the dependence on unobserved errors. This is required
    for the AICc model comparison and the model must pass the normality
    assumption for this estimation.

    $$
    L(\theta \mid y_1,\ldots,y_T),
    $$

5.  *Innovations Algorithm \| (MA)* : recursively computes optimal
    linear predictors and innovation variances, allowing estimation of
    MA parameters. It is efficient for pure MA models and handles
    missing data well.

6.  *Hannan-Rissanen Method \| (MA, ARMA)* : provides initial estimates
    for MA parameters, fits a higher order AR model to obtain
    pre-liminary residuals. This method regresses $Y_t$ on lagged values
    and estimated residuals and uses these as starting values for MLE.
    This method is widely used for ARMA models.

    |  |  |  |
    |------------------------|------------------------|------------------------|
    | **Model Type** | **Estimation Methods** | **Notes** |
    | AR (p) | Yule-Walker, Burg, CSS, MLE | Yule-Walker/Burg for initial estimates; MLE for refinement |
    | MA (q) | Innovations, Hannan-Rissanen, CSS, MLE | MA requires numerical methods; Hannan-Rissanen gives good starting values |
    | ARMA (p,q) | Hannan-Rissanen, CSS, MLE | Hannan-Risannen and MLE are standard |

    : *Summary of Parameter Estimation Methods*

**Overfitting and Parameter Redundancy**

**Principle of Model Selection: Keep it Simple**

### Model Estimation

### Model Diagnostics

Model diagnostics focus on assessing how well a model fits the observed
data and if the fit is inadequate, identifying suitable modifications.
The main objective is to evaluate the adequacy of the selected model.
After fitting the time series model, diagnostic checks on the residuals
should be performed to evaluate the model's fit. Residuals should behave
like white noise and should exhibit no autocorrelation, constant
variance, zero mean, approximate normality and no remaining seasonal
structure.

**Statistical Inference for Model Coefficients**

Each estimated parameter is evaluated for statistical significance. For
a generic coefficient $\beta$, the hypothesis test is:

$$
H_0:\ \beta = 0 \quad\text{(coefficient is not significant)}
$$

$$
H_1:\ \beta \neq 0 \quad\text{(coefficient is significant)}
$$

the test statistic is:

$$
z = \frac{\hat{\beta}}{SE(\hat{\beta})}
$$

and the corresponding p-value determines whether the coefficient is
significant. If the p-value is less than 0.05, we reject $H_0$ and the
coefficient is significant. Significant testing helps to identify
parameter redundancy and prevents overfitting by ensuring that only
meaningful terms are retained.

**Standardized Residuals:**

$$
e_t^{*} = \frac{\hat{e}_t}{\hat{\sigma}}
$$

are examined to assess whether the model errors exhibit: zero mean,
constant variance, no visible patterns, no remaining seasonality. A well
fitted model produces residuals that fluctuate randomly around zero with
no systematic structure.

**Normality Assessment:**

Approximate normality improves the reliability of forecast intervals.
Normality was evaluated using Normal Q-Q plot, Shapiro-Wilk test and a
histogram of standardized residuals. Extreme skewness or heavy tails may
indicate the need for further transformation or alternative modeling.

**Residual ACF and PACF:**

The residual ACF was examined to detect remaining serial dependence.
Significant spikes would indicate that the model failed to capture all
AR or MA structure. For an adequate model, all residual autocorrelations
lies within the 95% CI bounds, no seasonal spikes occur at lag and there
is no slow decay or cyclical pattern.

The residual PACF of the residuals are also inspected. Significant
spikes in the PACF suggest remaining autoregressive behavior. A well
specified model should show no significant partial autocorrelations and
no seasonal partial autocorrelations.

Together, the residual ACF and PACF confirm whether the model has
removed all linear dependence.

**Ljung-Box Test:**

The Ljung-Box Test formally evaluates whether the residuals are
uncorrelated:

$$
\begin{aligned}
H_0:\ &\text{Residuals are white noise} \\
H_1:\ &\text{Residuals are not white noise}
\end{aligned}
$$

A large p-value greater than 0.05 indicates failure to reject $H_0$
supporting the adequacy of the model.

### Forecasting

## Analysis and Results

### Data Preparation and Exploration

```{r, warning=FALSE, echo=T, message=FALSE}
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(readr)
library(dplyr)
library(zoo)
library(tseries)
library(TSA)

#install.packages("forecast")
library(forecast)

setwd("C:/Users/ayesh/OneDrive/Desktop/MS Data Science/IDC6940 Projects in Data Science/IDC6940_ARIMA")

```

```{r}
# Import Data Set
Data_Import <- read.csv("C:/Users/ayesh/OneDrive/Desktop/MS Data Science/IDC6940 Projects in Data Science/IDC6940_ARIMA/2025Q3 CMS HH MBI.csv")


# Remove N/A and Blank Values
Data <- Data_Import %>%
  filter(
    !is.na(`HH.Index.Levels`),          
    `HH.Index.Levels` != ""           
  )

# Convert HH Index Levels to Numeric
Data <- Data %>%
  mutate(`HH.Index.Levels` = as.numeric(`HH.Index.Levels`))

# Rename Headers

Data <- Data %>%
  rename(
    Date = `Market.Basket`,
    Index = `HH.Index.Levels`
  )

# Convert Date to Quarterly Format
Data <- Data %>%
  mutate(`Date` = as.yearqtr(Date, format = "%Y Q%q"))

# Sort Chronologically 
Data <- Data %>%
  arrange(`Date`)

# Convert to Time Series
TS_Data <- ts(Data$`Index`,
              start = c(2010,1),
              frequency = 4)

```

### Preliminary Analysis

```{r, warning=FALSE, echo=TRUE}

data(TS_Data)
head(TS_Data)
summary (TS_Data)

plot.ts(TS_Data, main = "CMS Home Health Market Basket Index 2010-2025")
qqnorm(TS_Data)
qqline(TS_Data, col= "red")
shapiro.test(TS_Data)
```

### Stationarity Test and Model Specification

```{r}
adf.test(TS_Data)
acf(as.numeric(TS_Data))
pacf(as.numeric(TS_Data))
eacf(as.numeric(TS_Data))
auto.arima(TS_Data)
```

Candidate Models

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
