---
title: "Forecasting CMS Market Basket Using Discrete Time Series Methods (ARIMA)"
author: "Ayesha Parsotan (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Introduction {.smaller}

Objective: Forecasting the Home Health market basket can support
financial planning, inform strategic decision‑making, and help
organizations anticipate implications for cost‑of‑care management. This
project applies classical time‑series methods to model and forecast the
CMS Home Health Market Basket Index.

Time‑series methods, particularly ARIMA models have been widely applied
in healthcare economics to forecast medical expenditures, inflation, and
service demand. Prior studies have demonstrated that ARIMA based
approaches perform well in modeling healthcare spending and medical
price behavior, including forecasting national health expenditures
@dritsakis2019time, providing broad overviews of health forecasting
methodologies @soyiri2013overview, and predicting medical service demand
using hybrid ARIMA‑based models @huang2020medical.

## What is a Time Series? {.smaller}

A Time Series is a collection of observations $x_t$ made sequentially in
time $T_0$. Studying models that incorporate dependence is the key
concept in time series analysis @cryer_chan_2010.

In general, the goal is to emphasize plotting methods that are
appropriate and useful for finding patterns that will lead to suitable
models for our time series data. Finding an appropriate model allows us
to then forecast future values based on past-behavior.

## What is a Time Series? {.smaller}

Discrete-Time Series & Continuous-Time Series

The CMS Home Health Market Basket is observed quarterly, meaning the
data occurs at fixed, evenly spaced intervals. Since the time series
advances in discrete steps, all models evaluated in this study are
formulated within a discrete time series stochastic process framework.

## Finding an Appropriate Model {.smaller}

There are three main steps in the process of finding an appropriate
model:

Model Specification (Identification)

Model Fitting

Model Diagnostics

## Stationarity and Time Series Foundations {.smaller}

A process $\{Y_t\}$ is strictly stationary if the joint distribution of
$(Y_{t_1}, \ldots, Y_{t_n})$ is the same as the joint distribution of
$(Y_{t_1-h}, \ldots, Y_{t_n-h})$ for all choices of time points
$t_1,\ldots,t_n$ and all choices of time lag $h$.

$$
f(Y_{t_1}, \ldots, Y_{t_n}) = f(Y_{t_1-h}, \ldots, Y_{t_n-h})
$$

## Weak Stationarity {.smaller}

A process $\{Y_t\}$ is weakly stationary if it satisfies that following
conditions:

$$ E(Y_t) = \mu $$ $$Var(Y_t) = \sigma^2$$

$$\gamma_{t,t-k} = \gamma_k$$

Weak stationarity is sufficient for ARMA and ARIMA models because they
assume that the underlying statistical process is stable over time.

## Non-Stationarity and White Noise {.smaller}

However, most data sets are non-Stationary: without constant mean over
time, and must be transformed. After removing systematic components of a
non-stationary series, the remaining series should resemble white noise:
a sequence of independent, identically distributed random variables.

```{r}
wn <- rnorm(500)  
plot.ts(wn, main = "White Noise Series")

```

## Trend, Seasonality, Periodicity and Heteroskedasticity {.smaller}

```{r}
par(mfrow =c(2,2))
data("uspop")
plot(uspop, ylab = "", yaxt = "n", main = "Trend")

data("lynx")
plot(lynx, ylab = "", yaxt = "n", main = "Periodicity/Seasonality")

data("JohnsonJohnson")
plot(JohnsonJohnson, ylab="",  yaxt = "n", main = "Heteroskedasticity")

data("AirPassengers")
plot(AirPassengers, ylab="",  yaxt = "n", main = "Heteroskedasticity, Seasonality, Trend")
```

## ADF Test {.smaller}

In practice, stationarity is assessed using visual diagnostics and
formal statistical tests before fitting time series models.

The Augmented Dickey-Fuller (ADF) test is used to assess stationarity.

$$
H_0: \text{The series is NOT stationary} \quad\text{vs.}\quad H_1: \text{The series is stationary}
$$

If the p-value is greater than 0.05, we will fail to reject $H_0$ and we
can use trend or seasonal estimation/elimination methods, transform the
times series data set and fit the non-stationary models.

## Models for Time-Series Analysis {.smaller}

Let $Y_t$ denote the observed times series, and $e_t$ represent an
unobserved white noise series. A general linear process, $Y_t$ is one
that can represented as a weighted linear combination of present and
past white noise terms as:

$$
Y_t = \mu + \sum_{j=0}^{\infty} \psi_j e_{t-j}
$$

where $\mu$ is a constant mean, $\{\psi_j\}$ is a sequence of
coefficients, and $\{e_t\}$ is a white-noise innovation process.

## Stationary Models {.smaller}

Moving Average (MA) process of order (q), MA(q)

$$Y_t = \mu - \theta_1 e_{t-1} - \cdots - \theta_q e_{t-q} + e_t$$

Autoregressive (AR) process of order (p), AR(p)

$$Y_t = \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p} + e_t$$

Autoregressive Moving Average (ARMA)

$$
Y_t
= \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p}
+ e_t
+ \theta_1 e_{t-1} + \cdots + \theta_q e_{t-q}.
$$

## Non-Stationary Models (ARIMA) {.smaller}

An Autoregressive Integrated Moving Average (ARIMA) model is an
extension of the ARMA model that incorporates non-seasonal differencing
or detrending to remove trend and achieve stationarity.

$$
\phi(B)(1-B)^d Y_t = \theta(B)e_t
$$

## Differencing {.smaller}

Differencing or de-trending transforms a non-stationary series into a
stationary one, allowing ARMA structure to be applied to the differenced
data. ARIMA models are appropriate when the series shows trend but no
seasonal pattern.

$$
\nabla^{d} X_t = (1 - B)^{d} X_t
$$

$$
\nabla X_t = X_t - X_{t-1}
$$

$$
\tilde{Y}_t = X_t - (\hat{\beta}_0 + \hat{\beta}_1 t)
$$

## Non-Stationary Models (SARIMA) {.smaller}

Seasonal Autoregressive Integrated Moving Average: SARIMA(p,d,q)(P,D,Q)s
model includes both nonseasonal components (p,d,q) and seasonal
components (P,D,Q) with seasonal period s. A time series ${Yt}$ follows
a SARIMA(p,d,q)(P,D,Q)s model if the differenced series is a stationary
ARMA (p,q) process with additional seasonal AR and MA terms.

$$
\phi(B)\,\Phi(B^s)\,(1-B)^d (1-B^s)^D Y_t
=
\theta(B)\,\Theta(B^s)\, e_t
$$

## Model Identification {.smaller}

In model specification, the classes of time series models are selected
that may be appropriate for a given observed series. A model chosen at
any time is tentative and subject to revision later on in the analysis.
In choosing a model, we should attempt to adhere to the principle of
parsimony; that is a model should require the smallest number of
parameters that will adequately represent the time series
@cryer_chan_2010.

## ACF, PACF, EACF {.smaller}

```{r}
ma1.1.s=arima.sim(list(order=c(0,0,1),ma=-0.9),100)
ma2.1.s=arima.sim(list(order=c(0,0,2),ma=c(-1,0.6)),100)
ar1.1.s=arima.sim(list(order=c(1,0,0),ar=0.9),100)
ar2.1.s=arima.sim(list(order=c(2,0,0),ar=c(1.5,-0.75)),100)
arma1.1.s=arima.sim(list(order=c(1,0,1),ar=0.6,ma=-0.3),100)

library(TSA)
par(mfrow = c(2, 3), oma = c(0, 0, 3, 0), mar = c(4, 4, 4, 1))
acf(ma1.1.s, main = "MA(1)")
acf(ma2.1.s, main = "MA(2)")
acf(ar1.1.s, main = "AR(1)")
pacf(ar1.1.s, main = "PACF AR(1)")
acf(ar2.1.s, main = "AR(2)")
pacf(ar2.1.s, main = "PACF AR(2)")


eacf(arma1.1.s)
```

## AICc {.smaller}

After candidate models are manually identified, they are compared using
the Corrected Akaike Information Criterion (AICc):

$$
AIC_c = AIC + \frac{2k(k+1)}{n - k - 1}
$$

where:

$$
AIC = -2\log(L) + 2k
$$

$L$ is the maximized likelihood,$k$ is the number of estimated
parameters, and$n$ is the sample size.

AICc adjusts for small sample sized and prevents overfitting. A lower
AICc equates to a better models, differences greater than 2 indicate
meaningful improvement and the most adequate model is selected. In
addition to manual identification using ACF, PACF, EACF, this study uses
auto.arima() function from the forecast package to support model
selection.

## Model Fitting (Parameter Estimation) {.smaller}


## 


## References
